---
title: Getting started with rWCVP
author: "Matilda Brown"
date: "`r format(Sys.time(), '%d %B, %Y')`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Getting started with rWCVP}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
knitr::opts_chunk$set(message = FALSE)
knitr::opts_chunk$set(warning = FALSE)
```

```{r}
library(rWCVP)
```

This R package has been designed to facilitate use of the World Checklist of Vascular Plants [more text here on how good it is etc also link to data paper]

# Distribution maps

We may want to generate a range map for a species, genus or family, like from POWO. 

```{r}
p <- plot_distribution(get_distribution("Callitris", rank="genus"))
```


```{r}
p #global
```
  




We can filter this map to only the native range by using introduced=FALSE (there are also location_doubtful and extinct that can be switched on and off).
```{r}
d <- get_distribution("Callitris", rank="genus", introduced=FALSE)
```
:::: {style="display: grid; grid-template-columns: 1fr 1fr; grid-column-gap: 10px;"}

::: {}


```{r}
p <- plot_distribution(d) #global
```
  
:::

::: {}

```{r}
p <- plot_distribution(d, crop.map = TRUE) #zoomed in
```

:::

::::

Here's a family example:
```{r}
p <- plot_distribution(get_distribution("Melastomataceae", rank="family"))
```

```{r}
p 
```
  
