---
title: "Mapping diversity with rWCVP"
author: "Matilda Brown"
date: "26/05/2022"
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  warning = FALSE,
  comment = "#>"
)
```


Here, I demonstrate how we can use some of the functions and data in the World Checklist to produce maps of various metrics (for distribution maps, see `plot_distribution`). 


The objective for this example is: **to generate maps showing various diversity metrics** (e.g. species richness, % endemism). 

In this example I use the pipe operator (`%>%`),`dplyr` and `ggplot` - if these are unfamiliar I'd suggest checking out https://tidyverse.org/ and some of the help pages therein, or this code might be difficult to interpret. Now, let's get started! 


```{r message=FALSE}
#load in some packages
library(rWCVP)
library(tidyverse) # for data manipulation
library(gt) # for tables
library(sf) #for spatial
```



### Species richness

The obvious place to start is global species richness. We can use `summary_table` to condense the global occurrence data for all species into raw counts per WGSRPD Level 3 Area. 

This gives us a list, with information in the first five slots and the data in the slot named `$Summary`. It's this `$Summary` dataframe that we'll be working with. 

```{r}
global_summary <- summary_table()
str(global_summary)
```
We could plot this using `summary_table_gt` but that's going to be a big table, so let's skip straight to plotting. `rWCVPdata` includes the area polygons for the WGSRPD Level 3 Areas - this is going to be the base that we add things to. Let's add the `global_summary` to the spatial data, and plot a map where species richness defines colour. 
```{r fig.asp=0.4822878}
#load the spatial data
area_polygons <- rWCVPdata::wgsprd3 %>% 
  #add the summary data, allowing for the different column names
  left_join(global_summary$Summary, by=c("LEVEL3_COD"="area_code_l3"))

ggplot(area_polygons)+
  geom_sf(aes(fill=Native))
```
<br>
Hmm, it's not as pretty as it could be and we're losing all the islands - let's make a few tweaks:
```{r fig.asp=0.43}
#wrapping n brackets so it assigns as well as prints
(p_native_richness <- ggplot(area_polygons) +
  #remove borders
  geom_sf(aes(fill=Native), col="transparent") +
  # add points for islands
  stat_sf_coordinates(aes(col=Native))+
  #remove borders
  theme_void()+
  #use a better colour palette
  scale_fill_viridis_c(name = "Native\nspecies")+ 
  scale_colour_viridis_c(name = "Native\nspecies")+ 
  #remove extra whitespace
  coord_sf(expand=FALSE)
)

  
```

Much better! 

<br> We can also incorporate other metrics - let's look at the proportion of species in each Level 3 Area that are endemic to that Area. We'll use a slightly different colour palette too, just for fun. 

```{r fig.asp=0.4822878}
area_polygons <- area_polygons %>% 
  mutate(percent_endemic = Endemic/Native)

(p_prop_endemic <- ggplot(area_polygons) +
  #remove borders
  geom_sf(aes(fill=percent_endemic), col="transparent") +
  # add points for islands
  stat_sf_coordinates(aes(col=percent_endemic))+
  #remove borders
  theme_void()+
  #use a better colour palette
  scale_fill_distiller(palette = "Reds",direction = 1, name = "% of species\nthat are\nendemic")+ 
  scale_colour_distiller(palette = "Reds",direction = 1,name = "% of species\nthat are\nendemic")+ 
  #remove extra whitespace
  coord_sf(expand=FALSE)
)
```

Endemism hotspots really jump out! We can use this information to produce a table of the top ten:
```{r}

area_polygons %>% 
  #get the region and continent names
  left_join(wgsrpd_mapping) %>% 
  #get the top 10
  slice_max(percent_endemic,n=10) %>% 
  #drop the spatial data
  st_drop_geometry() %>% 
  select(LEVEL3_NAM, percent_endemic, LEVEL1_NAM) %>% 
  group_by(LEVEL1_NAM) %>% 
  #format as a table
  gt() %>% 
  cols_label(
    percent_endemic = "% Endemic",
    LEVEL3_NAM = "WGSRPD Level 3 Area"
  ) %>% 
  tab_options(
    # some nice formatting
        column_labels.font.weight = "bold",
        row_group.font.weight = "bold",
        row_group.as_column = TRUE,
        data_row.padding = px(1),
        table.font.size = 12,
        table_body.hlines.color = "transparent",
        ) %>%
  #format the number as %
  fmt_percent(
    columns = percent_endemic,
    decimals = 1
  )
```
<br>
We can attach any variable to `area_polygons` - it doesn't have to be from the World Checklist! 
<br>
One more example, showing how we can limit our maps to a specific area of interest. Let's look at orchids in Africa. 


```{r}
orchid_africa_summary <- summary_table("Orchidaceae", "family", get_wgsrpd3_codes("Africa"))

```
Remember that the dataset is saved in the `$Summary` slot (trying to join `orchid_africa_summary` will cause an error). 

```{r }
#reset the spatial data by loading it in afresh
area_polygons <- rWCVPdata::wgsprd3 %>% 
  #add the summary data, allowing for the different column names
  left_join(orchid_africa_summary$Summary, by=c("LEVEL3_COD"="area_code_l3")) %>% 
  #add continent and region names for easy filtering
  left_join(wgsrpd_mapping) 

```

We could filter our polygons to only those in Africa, but from a mapping perspective it makes more sense to leave them in  (though greyed out). So, to constrain our map, we need to set up custom limits. 

```{r}
#get a sensible bounding box for the polygons in Africa
bounding_box <- st_bbox(area_polygons %>% filter(LEVEL1_NAM =="AFRICA"))
#add a buffer as we set limits
xmin <- bounding_box["xmin"] - 2
xmax <- bounding_box["xmax"] + 2
ymin <- bounding_box["ymin"] - 2
ymax <- bounding_box["ymax"] + 2

(p_orchids <- ggplot(area_polygons) +
  #remove borders
  geom_sf(aes(fill=Native), col="transparent") +
  # add points for islands
  stat_sf_coordinates(aes(col=Native), size=4)+
  #bounding box we set up above
  coord_sf(xlim=c(xmin, xmax), ylim=c(ymin, ymax))+
  #remove borders
  theme_void()+
  #set colour palette (fill for polygons, colour for points)
  scale_fill_viridis_c(name = "Native\nspecies")+ 
  scale_colour_viridis_c(name = "Native\nspecies")+
    ggtitle("Orchidaceae species richness")
)
```
 
For more examples and tutorials, see  {link to github page - doesn't exist yet. 
