---
title: "Publication-ready occurrence matrices with rWCVP"
author: "Matilda Brown"
date: "24/05/2022"
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  message = FALSE,
  warning = FALSE,
  comment = "#>"
)
```


In this example I use the pipe operator (`%>%`) and `dplyr` syntax - if these are unfamiliar I suggest checking out https://dplyr.tidyverse.org/ and some of the help pages therein. Now, let's get started! 

Load in some packages:

```{r}
library(rWCVP)
library(tidyverse) # for data manipulation
library(gt) # for tables
```
### Finding an example group
For this example, we don't have a particular area or group of plants that we want to examine, but this gives us a chance to showcase one of other the functions in `rWCVP`! We want a group of species that is a) not too large and b) distributed across a few WGSRPD Level 3 Areas. Brazil has good potential because it has five Level 3 Areas (a good number for this purpose because the table will fit on a portrait-oriented page). Let's see if there are some nice-sized example genera, using the `summary_table` function: 
```{r}
summary_table(taxon="Myrtaceae", rank="family", area=get_wgsrpd3_codes("Brazil"), 
              grouping.var = "genus") %>% summary_table_gt()
```
   <br> 
    
*Calycolpus* looks nice and tidy - let's see how the 8 species are distributed across the 5 areas. We can use the same function, but limit our taxon and change our grouping variable to area:
```{r}
summary_table(taxon="Calycolpus", rank="genus", area=get_wgsrpd3_codes("Brazil"), grouping.var = "area_code_l3") %>% summary_table_gt()
```
    
    
<br>
Hmm, maybe a bit too small - it only occurs in 3 of the 5 regions. What about *Myrciaria*? 

```{r}
summary_table(taxon="Myrciaria", rank="genus", area=get_wgsrpd3_codes("Brazil"), grouping.var = "area_code_l3") %>% summary_table_gt()

```
 <br>   
Perfect! 25 species (rows) won't take up too much space, and there are enough occurrences to make it interesting. 

### Generating and formatting the occurrence matrix
Now let's look at which species occur in which regions, using the `generate_occurence_matrix` function.  


```{r}
m <- generate_occurrence_matrix(taxon = "Myrciaria", rank = "genus",area = get_wgsrpd3_codes("Brazil"))
m
```
<br>
It's OK, but we can make it much prettier using the `gt` package.
Let's do the following: 

* remove the WCVP ID column
* change taxon_id to 'Species'
* make species names italic
* bold the column titles
* reduce the space around the text and make font size 12
* remove the internal borders
* change the 1s and 0s into X and blank


```{r}
m_gt <- m %>% 
  select(-plant_name_id) %>% #remove ID col
  gt() %>% 
  cols_label(
    taxon_name = "Species"
  ) %>% 
  #make species names italic
        tab_style(
        style=cell_text(style="italic"),
        locations = cells_body(
          columns= taxon_name
        )
      ) %>% 
  tab_options(
    # some nice formatting
        column_labels.font.weight = "bold",
        data_row.padding = px(1),
        table.font.size = 12,
        table_body.hlines.color = "transparent",
        ) %>%
  # change the zeroes into blanks
      text_transform(
        locations = cells_body(),
        fn = function(x){
          ifelse(x == 0, "", x)
        }
      ) %>% 
  # change the 1s into X
        text_transform(
        locations = cells_body(),
        fn = function(x){
          ifelse(x == 1, "X", x)
        }
      )
m_gt
```
<br>
Much nicer! We can save this `gt` table as an HTML table or as a picture. If we plan on making a few more tables, we can save space by saving our table style as a theme (see https://themockup.blog/posts/2020-09-26-functions-and-themes-for-gt-tables/ for more details on this)

```{r}
occ_mat_theme <- function(x){
  x %>% cols_label(
    taxon_name = "Species"
  ) %>% 
  #make species names italic
        tab_style(
        style=cell_text(style="italic"),
        locations = cells_body(
          columns= taxon_name
        )
      ) %>% 
  tab_options(
    # some nice formatting
        column_labels.font.weight = "bold",
        data_row.padding = px(1),
        table.font.size = 12,
        table_body.hlines.color = "transparent",
        ) %>%
  # change the zeroes into blanks
      text_transform(
        locations = cells_body(),
        fn = function(x){
          ifelse(x == 0, "", x)
        }
      ) %>% 
  # change the 1s into X
        text_transform(
        locations = cells_body(),
        fn = function(x){
          ifelse(x == 1, "X", x)
        }
      )
}
```
<br>
The biggest issue with gt() is that it doesn't support Word - for exporting directly to a docx file, check out `flextable` (https://ardata-fr.github.io/flextable-book/).
    
### Including or excluding occurrence types
What if we only want to know about native or introduced species? This function has the option to filter for one or the other. Brazilian *Myrciaria* doesn't look very interesting on that front (we can see from the summary table that only one species is introduced), so let's look at a more invasive group - *Poa* in Northern Europe (Level 2 Region). 

```{r}
summary_table(taxon="Poa", rank="genus", area=get_wgsrpd3_codes("Northern Europe"), 
              grouping.var = "area_code_l3") %>% summary_table_gt()
```
 <br>   
A few more there to work with. First, let's look at the native species only:
```{r}
m <- generate_occurrence_matrix(taxon = "Poa", rank = "genus",
                                area = get_wgsrpd3_codes("Northern Europe"), 
                                introduced=FALSE, extinct=FALSE, 
                                location_doubtful = FALSE)
m
```
   <br> 
We can format this matrix just like we did above, but let's skip that and go straight to introduced species only. We're doing all the same formatting as before, but also adding a heading - the `html` function makes it possiible to italicise our genus name and everything!

```{r}
m <- generate_occurrence_matrix(taxon = "Poa", rank = "genus",
                                area = get_wgsrpd3_codes("Northern Europe"), 
                                native=FALSE,
                                introduced=TRUE, extinct=FALSE, 
                                location_doubtful = FALSE)
m %>% 
  select(-plant_name_id) %>% #remove ID col
  gt() %>% 
  occ_mat_theme() %>%  #the theme we defined above
  #add a header
  tab_header(title=html("Introduced <em>Poa</em> species in Northern Europe")) 
```
  <br>  

### Bonus: adding a country spanner
Tables created with `gt` are extremely flexible - let's say we want to look at occurrences across the US-Canadian border:
```{r}
m <- generate_occurrence_matrix("Fritillaria", "genus", area = c("WAS", "ORE", "IDA","MNT", "ABT", "BRC"))


m_gt <- m %>% 
  select(-plant_name_id) %>% #remove ID col
  gt() %>% 
  occ_mat_theme() %>%  #the theme we defined above
  #add a header
  tab_header(title=html("<em>Fritillaria</em> species in Northwest USA and Southwest Canada")) 
m_gt
```
<br>
It would be really useful to know which of those codes are in the US and which are in Canada. We could use the data included in `rWCVP` to create a key:
```{r}
wgsrpd_mapping %>% 
  filter(LEVEL3_COD %in% c("WAS", "ORE", "IDA","MNT", "ABT", "BRC")) %>% 
  select(LEVEL3_NAM, LEVEL3_COD, COUNTRY) %>% 
  gt() %>% 
  #some formatting
  tab_options(
        column_labels.font.weight = "bold",
        data_row.padding = px(1),
        table.font.size = 12,
        table_body.hlines.color = "transparent",
        )
  
```
<br>
    
It would really be nicer to have it on the occurrence matrix though. Enter `tab_spanner()`:
```{r}
m_gt %>% 
  tab_spanner(label="United States",
              columns = c(IDA, MNT, ORE, WAS)
              ) %>% 
  tab_spanner(label="Canada",
              columns = c(ABT, BRC)
              ) 
```
<br>
There is a lot more that can be done with `gt` - see https://gt.rstudio.com/ for help, examples and documentation. 



